Generalized Parallel Computational Schemes\\ for Time-Consuming Global Optimization

### 2018-01-17 2017/KAZAN_LJM/paper_lobach.tex

%%%%%%%%%%%%%%%%%%%%%%% file template-ljm.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package ljm-auth
% for Lobachevskii Journal of Mathematics 2009/07/20
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[12pt]{amsart}

\usepackage{ljm-auth}

\usepackage{graphicx}

\usepackage{cite}

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}

\author{Strongin R.G., Gergel V.P., Barkalov K.A., Sysoyev A.V.}
\crauthor{Strongin, Gergel, Barkalov, Sysoyev} % for running heads

\tit{Generalized Parallel Computational Schemes for Time-Consuming Global Optimization}
\shorttit{Parallel Schemes for Global Optimization} %for running heads

\setcounter{page}{1}

\begin{document}

\maketit

\address{Lobachevsky State University of Nizhni Novgorod}

\email{gergel@unn.ru}

\abstract{This paper addresses computationally intensive global optimization problems, for solving of which the supercomputing systems with exaflops performance can be required. To overcome such computational complexity, the paper proposes the generalized parallel computational schemes, which may involve numerous efficient parallel algorithms of global optimization. The proposed schemes include various ways of multilevel decomposition of parallel computations to guarantee the computational efficiency of supercomputing systems with shared and distributed memory multiprocessors with thousands of processors to meet global optimization challenges.}

\notes{0}
{
%\subclass{} % 2010 Mathematical Subject Classification

\keywords{
global optimization, information-statistical theory, parallel computations, high-performance computing, supercomputing technologies
}

\thank{
This research was supported by the Russian Science Foundation, project No 16-11-10150 "Novel efficient methods and software tools for time-consuming decision making problems using supercomputers of superior performance"
}
}

@@

@@

@@

As it is well known (see, for example, \cite{Strongin1}), the numerical evaluation of the global optimum points of the minimized functions requires constructing a coverage of the search domain $D$. This results in a very high computational intensity of global optimization problems even in the case of small dimensionality.

The computational complexity can be reduced considerably if the computational grids resulting from covering the search domain are non-uniform, i.e. when the grid nodes are distributed denser around the global optimal point only. The construction of such non-uniform coverings may be ensured in the case of an adaptive computation, when in order to select the new trial points, all search information obtained in the course of computation (previous trial points and the values of the minimized function at these points) is used. This necessary condition complicates the computational schemes of the global optimization methods essentially as it involves the computationally intensive analysis of large amounts of the multidimensional search information. As a result, many optimization algorithms use various dimensionality reduction schemes to certain extent (see, for example, \cite{Strongin1, Strongin2, Sergeyev2}).

@@

@@

@@

@@

The information-statistical theory \cite{Strongin1, Strongin3} formed a basis for many efficient global optimization methods -- see, for example \cite{Strongin2, Sergeyev4, Gergel1, Gergel2, Barkalov, Gergel3, Gergel4, Gergel5, Lera}. Moreover, a general computation scheme \cite{Strongin1, Strongin2, Strongin3, Grishagin1, Grishagin2} was proposed for all developed global search algorithms. This scheme can be represented as follows.

@@

The computed characteristics $R(i), 1 \leq i \leq k + 1$ can be understood as the importance indicators for the intervals with respect to the global minimum point location. This provides the explanation for the interval selection scheme for executing the next trial: the point of every next global search iteration is selected in the interval with the maximum characteristic (i.e. in the interval, which is the most likely one to contain the global minimum point).

@@

As it has been noted above, the general approach to provide the parallel computations consists in simultaneous calculating the minimized function values at several different points of the searching domain $D$. As the interval characteristics $R(i), 1 \leq i \leq k + 1$ (see Subsection 3.1) are understood as the importance indicators for the intervals with respect to the global minimum point location, additional intervals for several trials to be run in parallel should be selected subject to the following values of the characteristics.

@@

In general, the described general scheme of parallel algorithms of global optimization for systems with shared memory identifies the procedure for competitive parallel computations (see Fig.~1a) where the search intervals $(x_{i-1} ,x_i), 1 \leq i \leq k + 1$ compete with each other for the access to the processors to execute next global search iterations. Taking into account the possible number of processors/cores in computational nodes with homogeneous shared memory, the described approach enables the parallelism of about $10^2$ computing units.

\begin{figure}[t]
\centering
\includegraphics[width=13.0cm]{Fig1}
\caption{Generalized computational scheme of parallel algorithms}
\label{fig:Fig1}
\end{figure}

The conditions of convergence and nonredundancy of the resulting parallel computations for the global optimization algorithms formulated in the framework of general scheme of parallel algorithms of global optimization are given in \cite{Strongin1, Strongin2, Strongin4, Grishagin2}.

@@

Using such a parallelization approach can result in the following general scheme of parallel algorithms of global optimization for the systems with distributed and shared memory -- see also \cite{Gergel1} and Fig.~1b.

@@

@@

To enable the parallel computations, numerous nested optimization subproblems can be generated at once on each decomposition level for solving these ones in parallel \cite{Sergeyev7} (see Fig.~1c). The resulting number of parallel subproblems can be controlled using a predefined parallelization vector
\begin{equation}
\pi = (\pi_1, \pi_2, ..., \pi_M),
\end{equation}
where $\pi_i, 1 \leq i \leq M$, is the number of subproblems in the $(i+1)$-th decomposition level being solved in parallel. For the $M$-th level, $\pi_M$ means the number of parallel trials in the course of minimization of $\varphi_M(u_1,...,u_M) = \varphi_M(y_1,...,y_N)$ by the variable $u_M$  subject to the fixed values of $u_1, \dots, u_{M-1}$. Thus, the total number of the involved processors/cores equals to
\begin{equation}
\Pi = 1 + \sum_{i=1}^{M-1} \prod_{j=1}^i \pi_j.
\end{equation}

@@

To obtain more reliable estimates of the parallelization efficiency, 100 test problems of global optimization were solved. The maximum allowed number of parallel iterations was $Kmax = 10\,000$.

@@

The paper proposes a generalized computational scheme of parallel algorithms for solving global optimization problems, which require the exascale computations. The developed approach includes a competitive parallel computational scheme for the systems with shared memory, the distributed parallel computations based on using the multiple Peano curves for the dimensionality reduction, and the multilevel decomposition of the parallel computations for the exascale supercomputers. In addition, the supercomputers may include also Intel Xeon Phi or GPGPU-based accelerators. The proposed schemes ensure the computational efficiency of the supercomputing systems with shared and distributed memory with thousands of processors to address the global optimization challenges.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

\documentclass[
11pt,%
tightenlines,%
twoside,%
onecolumn,%
nofloats,%
nobibnotes,%
nofootinbib,%
superscriptaddress,%
noshowpacs,%
centertags]%
{revtex4}
\usepackage{ljm}

\setcounter{page}{3}

\begin{document}
\titlerunning{Generalized Parallel Schemes for Global Optimization}
\authorrunning{Strongin, Gergel, Barkalov, Sysoyev}

\title{Generalized Parallel Computational Schemes\\
for Time-Consuming Global Optimization}

\author{\firstname{R.~G.}~\surname{Strongin}}
\email[E-mail: ]{president@unn.ru}
\affiliation{Lobachevsky State University of Nizhni Novgorod}

\author{\firstname{V.~P.}~\surname{Gergel}}
\email[E-mail: ]{gergel@unn.ru}
\affiliation{Lobachevsky State University of Nizhni Novgorod}

\author{\firstname{K.~A.}~\surname{Barkalov}}
\email[E-mail: ]{barkalov@vmk.unn.ru}
\affiliation{Lobachevsky State University of Nizhni Novgorod}

\author{\firstname{A.~V.}~\surname{Sysoyev}}
\email[E-mail: ]{sysoyev@vmk.unn.ru}
\affiliation{Lobachevsky State University of Nizhni Novgorod}

\firstcollaboration{(Submitted by ) }

\received{}

\begin{abstract}
This paper addresses computationally intensive global optimization problems, for solving of which the supercomputing systems with exaflops performance can be required. To overcome such computational complexity, the paper proposes the generalized parallel computational schemes, which may involve numerous efficient parallel algorithms of global optimization. The proposed schemes include various ways of multilevel decomposition of parallel computations to guarantee the computational efficiency of supercomputing systems with shared and distributed memory multiprocessors with thousands of processors to meet global optimization challenges.
\end{abstract}

\subclass{90C26, 90C30}

\keywords{global optimization, information-statistical theory, parallel computations, high-performance computing, supercomputing technologies}

\maketitle

@@

From all the above, it follows that solving the global optimization problems within reasonable times is possible only utilizing the exascale supercomputers, which use the high-performance parallel algorithms of global optimization

@@

The numerical solution of a global optimization problem consists in finding a numerical estimate $\bar{y}^\ast$ of the global minimum point $y^\ast \in D$ with a required accuracy $\varepsilon > 0$, i.e.

\[
\|\bar{y}^\ast - y^\ast\| \leq \varepsilon.
\]

@@

As it is well known (see, for example, \cite{Strongin1}), the numerical evaluation of the global optimum points of the minimized functions requires constructing a coverage of the search domain $D$. This results in a very high computational intensity of global optimization problems even in the case of small dimensionality. For example, if a uniform grid is used to cover the search domain, the number of grid nodes increases exponentially with increasing problem dimensionality, i.e. 
\[
K = K(\varepsilon) \approx \prod_{i=1}^N[(b_i - a_i) / \varepsilon],
\]
where $\varepsilon > 0$ is the uniform grid step (i.e. even for a grid with $10$ nodes per axis and $N = 20$, the total number of nodes is as large as $10^{20}$).

The computational complexity can be reduced considerably if the computational grids resulting from covering the search domain are non-uniform, i.e. when the grid nodes are distributed denser around the global optimal point only (see Fig.~1). The construction of such non-uniform coverings may be ensured in the case of an adaptive computation, when in order to select the new trial points, all search information obtained in the course of computation (previous trial points and the values of the minimized function at these points) is used. This necessary condition complicates the computational schemes of the global optimization methods essentially as it involves the computationally intensive analysis of large amounts of the multidimensional search information. As a result, many optimization algorithms use various dimensionality reduction schemes to certain extent (see, for example, \cite{Strongin1, Strongin2, Sergeyev2}).

\begin{figure}
\centering
\includegraphics[height=5.0cm]{Fig1a}
\includegraphics[height=5.0cm]{Fig1b}
\caption{Uniform and non-uniform grid}
\label{fig:Fig1}
\end{figure}

@@

Fig.~2 shows an approximation of the Peano curve mapping a two-dimensional domain with the third mapping density level. The accuracy of the Peano curve approximation is determined by its construction density level. The order of mapping the two-dimensional domain $D$ is determined by the subdomain connection order resulting from the construction of the Peano curve approximation; the corresponding order of subintervals of the interval $[0, 1]$ is determined by their ordered placement from left to right (i.e. the subdomain $\Delta_1$ corresponds to the subinterval $\delta_1$, the subdomain $\Delta_2$ corresponds to the subinterval $\delta_2$, etc).

\begin{figure}
\centering
\includegraphics[height=5.0cm]{Fig2}
\caption{An example of the Peano curve approximation with the third density level}
\label{fig:Fig2}
\end{figure}

It should be noted that reducing the multidimensional functions from (1) to the corresponding one-dimensional functions (3) using Peano-type curves preserves the uniform boundedness of the function value differences subject to the bounded argument variations, i. e. ensures the feasibility of the condition (4). Hence the computationally intensive analysis of the multidimensional subdomains to construct a non-uniform coverage of the search domain D can be replaced by a more simplified analysis of reduced one-dimensional data.

@@

A key distinctive feature of the global optimization problems is a limited opportunity to use parallelization methods used widely in the theory and practice or high-performance computing.
\begin{itemize}
\item \textit{Data parallelism} (distributing the computational domain among different computing units) is inefficient as in this case only one processor will process the subdomain that contains the sought global minimum while other processors will perform the redundant computations.

\item \textit{Task parallelism} (parallelization of computational schemes) is also inefficient as computational costs of the execution of the optimization algorithm schemes are relatively low (in accordance with the initial supposition, the most time-consuming calculations in the course of global optimization consist in computing the optimized function values).

\item \textit{Parallelization of time-consuming computations} to compute the optimized function values is not general (parallelization methods have to be developed from scratch every time a new optimization problem is solved).
\end{itemize}

@@

It should be noted that the proposed approach can also be applied for solving many other computationally intensive problems -- such as root finding in systems of nonlinear equations, integration of multidimensional integrals, approximation of the data dependencies etc.

In accordance with the approach mentioned above, in this paper a generalized computational scheme of parallel computations to solve the global optimization problems are described as follows. To start with, in Subsection 3.1, the general computation scheme of global search algorithms is given. Then, in Subsection 3.2, the general scheme of parallel computations for the systems with shared memory is presented. In Subsection 3.3, a parallel computational scheme for the systems with distributed memory is described. Finally, in Subsection 3.4, a multilevel decomposition scheme for parallel computations for solving the global optimization problems is proposed.

@@

The information-statistical theory \cite{Strongin1, Strongin3} formed a basis for many efficient global optimization methods -- see, for example \cite{Strongin2, Sergeyev4, Gergel1, Gergel2, Barkalov, Gergel3, Gergel4, Gergel5, Lera}. Moreover, a general computation scheme \cite{Strongin1, Strongin2, Strongin3, Grishagin1, Grishagin2} (see also Fig.~3) was proposed for all developed global search algorithms. This scheme can be represented as follows.

\begin{figure}
\centering
\includegraphics[height=6.0cm]{Fig3}
\caption{General computation scheme of global optimization algorithms}
\label{fig:Fig3}
\end{figure}

@@

In order to explain the computational scheme given above, the following can be noted. The computed characteristics $R(i), 1 \leq i \leq k + 1$ can be understood as the importance indicators for the intervals with respect to the global minimum point location. This provides the explanation for the interval selection scheme for executing the next trial: the point of every next global search iteration is selected in the interval with the maximum characteristic (i.e. in the interval, which is the most likely one to contain the global minimum point).

@@

As it has been noted above, the general approach to provide the parallel computations consists in simultaneous calculating the minimized function values at several different points of the searching domain $D$. For the shared memory computing systems, selecting the set of points for executing the parallel trials can be based on the characteristics $R(i), 1 \leq i \leq k + 1$ of the search intervals $(x_{i-1} ,x_i), 1 \leq i \leq k + 1$ introduced in the framework of the general computational scheme (see Subsection 3.1). As the characteristics are understood as the importance indicators for the intervals with respect to the global minimum point location, additional intervals for several trials to be run in parallel should be selected subject to the following values of the characteristics.

@@

In general, the described general scheme of parallel algorithms of global optimization for systems with shared memory identifies the procedure for competitive parallel computations (see Fig.~4a) where the search intervals $(x_{i-1} ,x_i), 1 \leq i \leq k + 1$ compete with each other for the access to the processors to execute next global search iterations. Taking into account the possible number of processors/cores in computational nodes with homogeneous shared memory, the described approach enables the parallelism of about $10^2$ computing units.

\begin{figure}[t]
\centering
\includegraphics[width=13.0cm]{Fig4}
\caption{Generalized computational scheme of parallel algorithms}
\label{fig:Fig4}
\end{figure}

The conditions of convergence and nonredundancy of the resulting parallel computations for the global optimization algorithms formulated in the framework of general scheme of parallel algorithms of global optimization are given in \cite{Strongin1, Strongin2, Strongin4, Grishagin2}. Thus, in particular, when the convergence conditions are met, the parallel computing for the MGAGS algorithm is non-redundant involving up to $2^N$ processors (cores) where $N$ is the dimension of the global optimization problem to be solved.

@@

Using such a parallelization approach can result in the following general scheme of parallel algorithms of global optimization for the systems with distributed and shared memory -- see also \cite{Gergel1} and Fig.~4b.

@@

It should be noted that this parallel computing scheme does not have any central control node. The number of computational nodes may vary during the global optimization, and removing any node will not lead to the loss in finding the global minimum of the optimization problem (1). In addition, the data communication may be asynchronous (the computational nodes should not wait until the optimization data are transmitted).

@@

To enable the parallel computations, numerous nested optimization subproblems can be generated at once on each decomposition level for solving these ones in parallel \cite{Sergeyev7} (see Fig.~4c). The resulting number of parallel subproblems can be controlled using a predefined parallelization vector
\begin{equation}
\pi = (\pi_1, \pi_2, ..., \pi_M),
\end{equation}
where $\pi_i, 1 \leq i \leq M$, is the number of subproblems in the $(i+1)$-th decomposition level being solved in parallel. For the $M$-th level, $\pi_M$ means the number of parallel trials in the course of minimization of $\varphi_M(u_1,...,u_M) = \varphi_M(y_1,...,y_N)$ by the variable $u_M$  subject to the fixed values of $u_1, \dots, u_{M-1}$. Thus, the total number of the involved processors/cores equals to
\begin{equation}
\Pi = 1 + \sum_{i=1}^{M-1} \prod_{j=1}^i \pi_j.
\end{equation}

@@

To obtain more reliable estimates of the parallelization efficiency, 100 test problems of global optimization were solved. The global minima $y^\ast$ of the test problems were considered to be found if a trial point $y^k$ is generated in the $\delta$-vicinity of the global minimum, i.e. $\| y^k - y^ \ast \| \leq \delta$. The vicinity size was set equal to $\delta = 0.01 \|b - a \|$ , where $a$ and $b$ were the boundaries of the search domain $D$. The maximum allowed number of parallel iterations was $Kmax = 10\,000$.

@@

The paper proposes a generalized computational scheme of parallel algorithms for solving global optimization problems, which require the exascale computations. The developed approach includes a competitive parallel computational scheme for the systems with shared memory, the distributed parallel computations based on using the multiple Peano curves for the dimensionality reduction, and the multilevel decomposition of the parallel computations for the exascale supercomputers. In addition, the supercomputers may include also Intel Xeon Phi or GPGPU-based accelerators. The proposed schemes ensure the computational efficiency of the supercomputing systems with shared and distributed memory with thousands of processors to address the global optimization challenges. The executed computational experiments confirm the potential of the proposed approach.

@@

\begin{acknowledgments}
This research was supported by the Russian Science Foundation, project No 16-11-10150 "Novel efficient methods and software tools for time-consuming decision making problems using supercomputers of superior performance".
\end{acknowledgments}
